\section{08 \textbar\ Toward Self-Hosting RCX}

Up to now, RCX-$\pi$ has been evaluated by an \emph{external Python engine}.
The system reduces motifs, but the reducer itself is not yet a motif.
Self-hosting begins when the evaluator becomes representable \emph{inside} RCX.

\[
\text{RCX-$\pi$ runs motifs.}
\qquad
\text{RCX aims for motifs that run RCX.}
\]

This chapter defines the scaffolding needed to cross that threshold.

%============================================================
\subsection{Philosophical Boundary}

A self-hosted evaluator must satisfy:

\begin{enumerate}
\item The evaluator is expressed as a motif.
\item A motif can apply reduction rules to another motif.
\item The system can evaluate \emph{itself}.
\item The host evaluator only bootstraps once.
\end{enumerate}

The Python evaluator is scaffolding ---
a ladder we will eventually remove.

\[
\text{goal: } \mu(\text{eval}, M) \Rightarrow M'
\]

no bytecode, no IR, no separate runtime.

Only motifs.

%============================================================
\subsection{Representation of Rules as Structure}

Rewrite rules become first-class objects.

A rule becomes:

\[
\text{rule} := \mu(\text{pattern},\text{replacement})
\]

For example:

\[
\text{swap}(x,y) \Rightarrow (y,x)
\quad\mapsto\quad
R_{\text{swap}} = \mu(\mu(\text{swap},\mu(x,y)), \mu(y,x))
\]

The evaluator motif is then a set of rules plus a reduction driver:

\[
\text{EVAL} := \mu(R_1,R_2,\dots,R_n)
\]

Later, RCX-π could evolve rules dynamically.
Structure chooses structure.

%------------------------------------------------------------
\subsection{The Self-Host Scaffold}

In \texttt{self\_host.py}, we introduce the seed evaluator:

\[
\text{eval\_seed}(M)
   := \text{search for matching rule in } EVAL
\]

The steps:

\begin{enumerate}
\item Encode rules as motifs
\item Encode matching logic as motifs
\item Encode single-step reduction as motif
\item Construct \texttt{eval\_seed} with no Python logic
\item Run RCX inside RCX
\end{enumerate}

At first, matching may be external.
Full replacement will be internal later.

%------------------------------------------------------------
\subsection{Safety and Divergence}

Self-hosting introduces risk:

\[
\mu(\text{eval},\text{eval}) \Rightarrow ?
\]

Unrestricted recursors can explode geometrically.
We introduce a \emph{safe reduction gate}:

\[
\text{safe}(M) =
\begin{cases}
M & \text{if pure or structural} \\
\text{blocked} & \text{if meta-tagged or cyclic}
\end{cases}
\]

detected via:

\begin{itemize}
\item purity tests
\item meta-tags
\item cyclic shape trace
\end{itemize}

The goal is not to prevent complexity,
but to prevent infinite blind rewriting.

\[
\text{chaos is allowed, but understood.}
\]

%------------------------------------------------------------
\subsection{Proof-of-Concept: RCX Evaluating RCX}

A minimal working dream-demo:

\[
\mu(\text{EVAL},\mu(\text{swap},(2,5)))
   \Rightarrow (5,2)
\]

Everything—program, data, evaluator—inside the same universe.

This is the first spark of \emph{RCX proper}.

%------------------------------------------------------------
\subsection{Roadmap to Full Self-Hosting}

\begin{enumerate}
\item Encode evaluator rules structurally
\item Add pattern-matcher motif
\item Add rewrite motif
\item Store evaluator as motif
\item Evaluate arbitrary motifs internally
\item Remove Python dependency
\end{enumerate}

Optional future extensions:

\begin{itemize}
\item partial evaluation
\item speculative reduction
\item hyper-programs (programs rewriting programs)
\item lobe-membranes for space-bounded execution
\end{itemize}

RCX becomes an environment, not a library.

%------------------------------------------------------------
\subsection{Status Today}

\begin{itemize}
\item Core evaluator implemented externally ✓
\item Programs encoded as motifs ✓
\item Higher ops, pretty-printer, REPL ✓
\item Self-host seed exists (experimental) ~
\item Full evaluator motif pending ✗
\end{itemize}

We now stand at the boundary where the scaffold can be replaced
piece by piece by structural equivalents.

\[
\textbf{RCX learns to run itself.}
\]

%------------------------------------------------------------
\subsection{Next: Growth, Evolution, and Emergence}

Chapter 09 will focus on:

\begin{itemize}
\item evolving rule-sets
\item program-mutation as computation
\item reflective meta-evaluation
\item RCX organisms (lobe growth + hydration loops)
\end{itemize}

The future of RCX is not a language,
but a \emph{living reduction environment}.

\[
\textbf{09 \textbar\ Emergent RCX and Evolution Engines}
\]